---
title: "Corridor edges"
format: html
---

```{r}
#| label: setup
#| message: false

library("dplyr")
library("leaflet")
library("lwgeom")
library("osmdata")
library("purrr")
library("sf")
library("sfnetworks")
library("tidygraph")
```

In this notebooks we explore how to delineate river corridor edges using Berlin as the study area. We focus on the Seine and use a specific projected CRS for the analysis.

```{r}
city_name <- "Berlin"
river_name <- "Spree"
epsg_code <- 32633  # UTM zone 33N
bbox_buffer <- 2000
```

We start by getting the bounding box for the study area.

```{r}
# bounding box
bb <- getbb(city_name)
bbox <- bb |> as.vector()
names(bbox) <- c("xmin", "ymin", "xmax", "ymax")
bbox <- st_bbox(bbox, crs = st_crs(4326)) |>
    st_as_sfc() |>
    st_transform(epsg_code)
bbox_expanded <- bbox |>
    st_buffer(bbox_buffer)
```

A couple of utility functions:

```{r}
# query the Overpass API for a key:value pair within a given bounding box
osmdata_as_sf <- function(key, value, bb){
    bb |>
        opq() |>
        add_osm_feature(key = key, value = value) |>
        osmdata_sf()
}
```

```{r}
# get geometry in lat/lon (WGS84)
getGeomLatLon <- function(x) st_transform(x, 4326) |> st_geometry()
```
```{r}
# get city boundary
city_boundary <- osmdata_as_sf("place", "city", bb)

city_boundary <- city_boundary$osm_multipolygons |>
   st_geometry() |> st_transform(epsg_code)

city_boundary <- city_boundary[2] # consider only Berlin
```
```{r}
leaflet() |>
    addTiles() |>
    addPolygons(data = getGeomLatLon(bbox_expanded), color = "red") |>
    addPolygons(data = getGeomLatLon(bbox), color="blue") |>
    addPolygons(data = getGeomLatLon(city_boundary), color="red")

```

The Spree flows into another river to the West of Berlin - limit bbox, also to the East:
```{r}
bbox <- c(13.25, 52.33824, 13.55, 52.67551)
names(bbox) <- c("xmin", "ymin", "xmax", "ymax")
bbox <- st_bbox(bbox, crs = st_crs(4326)) |>
    st_as_sfc() |>
    st_transform(epsg_code)
bbox_expanded <- bbox |>
    st_buffer(bbox_buffer)
```
```{r}
leaflet() |>
    addTiles() |>
    addPolygons(data = getGeomLatLon(bbox_expanded), color = "red") |>
    addPolygons(data = getGeomLatLon(bbox), color="blue")
```


## 1. Initial corridor edge ----

While ideally we want to base the initial estimate of the corridor edge on the basis of the river valley delineation, we use here the roughest approach of defining a buffer region around the waterways. This method could actually be the method of choice for urban areas with flat topographies.

### Waterways

Querying the Overpass API for `waterway:river`:

```{r}
# waterways (linestrings)
waterways <- osmdata_as_sf("waterway", "river", bb)
```

OSM multilines include river lines grouped by the river name. We extract the relevant waterway and transform to the projected CRS:

```{r}
waterway <- waterways$osm_multilines |>
    filter(name == river_name) |>
    st_transform(epsg_code) |>
    st_geometry() |>
    st_intersection(st_buffer(bbox, bbox_buffer + 1000))

# st_write(waterway, "../data/generated/waterway.gpkg")
```

```{r}
leaflet() |>
    addTiles() |>
    addPolylines(data = getGeomLatLon(waterway), color="blue") |>
    addPolygons(data = getGeomLatLon(bbox), color="red")
```

### Water

Querying the Overpass API for `natural:water`:

```{r}
# water area (polygons)
water <- osmdata_as_sf("natural", "water", bb)
```

The results of the query with tag `natural:water` also include features such as fountains. The geometries are not contiguous and some part of the water bodies are actually represented as lines instead of polygons:

```{r}
leaflet() |>
    addTiles() |>
    addPolylines(data = getGeomLatLon(water$osm_lines), color="blue", group="osm_lines") |>
    addPolygons(data = getGeomLatLon(water$osm_polygons), color="red", group="osm_polygons") |>
    addPolygons(data = getGeomLatLon(water$osm_multipolygons), color="black", group="osm_multipolygons") |>
    addLayersControl(overlayGroups=c("osm_lines", "osm_polygons", "osm_multipolygons"))
```

We determine which features intersect the relevant waterway:

```{r}
waterbody <- bind_rows(water$osm_polygons, water$osm_multipolygons) |>
    st_transform(epsg_code) |>
    st_filter(waterway, .predicate = st_intersects) |>
    st_geometry() |>
    st_union()
```

```{r}
leaflet() |>
    addTiles() |>
    addPolylines(data = getGeomLatLon(waterway), color="blue") |>
    addPolygons(data = getGeomLatLon(waterbody), color="red")
```

### Constructing the initial corridor

We construct the initial corridor from the merged geometries of `waterway` and `waterbody`, with a buffer region around them:

```{r}
buffer_dist <- 500  # distance (in m) from the water stream
```

```{r}
corridor_initial <- c(waterway, waterbody) |>
    st_buffer(buffer_dist) |>
    st_union()
```

```{r}
leaflet() |>
    addTiles() |>
    addPolygons(data = getGeomLatLon(corridor_initial))
```

```{r}
# save the waterway geometries (string and polygon)
st_write(c(waterway, waterbody), dsn = sprintf("data/generated/waterway_%s.gpkg", river_name))
```

## 2. Street network ----

Querying the Overpass API for the `highway` key:

```{r}
highways_value <- c("motorway", "primary", "secondary", "tertiary")
highways <- osmdata_as_sf("highway", highways_value, bb)

# route_value <- c("road")
# route <- osmdata_as_sf("route", route_value, bb)
```

We use `sfnetworks` to setup the street network based on the OSM data, mostly following [this tutorial](https://geospatial-community.netlify.app/post/2022-03-31-spatial-networks/) for data cleaning and network setup.

```{r}
# cast polygons (closed streets) into lines
poly_to_lines <- highways$osm_polygons |>
    st_cast("LINESTRING")
# # include road segments found as route:road in OSM
# routes <- route$osm_lines |>
#     st_intersection(st_transform(bbox_expanded, 4326)) |>
#     filter(highway %in% highways_value) |>
#     st_cast("LINESTRING")
# combine all features in one data frame
highways_lines <- highways$osm_lines |>
    bind_rows(poly_to_lines)
    # \> bind_rows(routes)

# st_write(highways_lines, "../data/generated/streets.gpkg")
```

```{r}
# create network, only keeping "highway" column
net <- highways_lines |>
    select("highway") |>
    as_sfnetwork(directed = FALSE) |>
    st_transform(epsg_code)
```

```{r}
getNodes <- function(net) net |> activate("nodes") |> st_geometry()
getEdges <- function(net) net |> activate("edges") |> st_geometry()

leaflet() |>
    addTiles() |>
    addPolylines(data = getEdges(net) |> getGeomLatLon(), color = "black") |>
    addCircles(data = getNodes(net) |> getGeomLatLon(), color = "red") |>
    addPolygons(data = getGeomLatLon(corridor_initial), color = "blue")
```

```{r}
# save the street network
edges <- net |> st_as_sf("edges")
nodes <- net |> st_as_sf("nodes")

st_write(edges, dsn = sprintf("data/generated/street_network_edges_%s.gpkg", city_name), append = FALSE)
st_write(nodes, dsn = sprintf("data/generated/street_network_nodes_%s.gpkg", city_name), append = FALSE)
```

## 3. Corridor edge delineation

Define the area of interest (AoI), and split it in two parts using the waterway as separator. The two areas can then be used to :

```{r}
# split the AoI using the waterway
areas <- bbox_expanded |>
    st_split(waterway) |>
    st_collection_extract()
```

Actually the area can be fragmented into multiple arease:
```{r}
leaflet() |>
    addTiles() |>
    addPolylines(data = areas |> getGeomLatLon(), color = "blue")
```

Determine the "vertices" of the initial river corridor as the intersections of the initial river corridor with the AoI boundary. We will use these points as extremes for the corridor edges:

```{r}
vertices <- bbox |>
    st_boundary() |>
    st_intersection(corridor_initial) |>
    # this should consists of two linestring components, determine the endpoints
    st_cast("POINT")
```
ik8
Consider only one side of the river for now (ideally we will do a mapping for all the steps below to the elements of `areas`)

```{r}
area <- areas[1]
```

Trim the street network on the basis of the selected area, also dropping all nodes within the initial corridor. This approach might actually be too "strict", since it makes it impossible for the corridor edges to cross the initially-defined corridor lines (e.g. via the valley edge). An alternative approach that we might pursue could be to define a cost proportional to the distance from the initial corridor edges, to be added to the "weight" of the network edges before looking for the shortest path. The additional cost would ideally try to keep the path as close as possible to the initial corridor edges, while making it possible for the ultimate edge to cross the initial boundaries.

```{r}
not_intersects <- function(x, y) !st_intersects(x, y)

trim <- function(net, area, corridor_initial){
    net |>
        activate("nodes") |>
        # select all nodes in the given areas
        st_filter(area, .predicate = st_intersects) |>
        # drop all nodes in the initial corridor
        st_filter(corridor_initial, .predicate = not_intersects)
}
```

```{r}
trimmed <- trim(net, area, corridor_initial)
```

We perform standard cleaning tasks on the graph:

```{r}
#' Simplify the graph
#'
#' Remove loops and double-edge connections
#' https://luukvdmeer.github.io/sfnetworks/articles/sfn02_preprocess_clean.html#simplify-network
simplify <- function(net){
    net|>
        activate("edges") |>
        # reorder the edges so that the shortest is kept
        arrange(edge_length()) |>
        filter(!edge_is_multiple()) |>
        filter(!edge_is_loop())
}


clean <- function(net){
    net |>
        simplify() |>
        # subdivide edges by adding missing nodes
        # https://luukvdmeer.github.io/sfnetworks/articles/sfn02_preprocess_clean.html#subdivide-edges
        convert(to_spatial_subdivision, .clean = TRUE) |>
        # remove pseudo-nodes
        # https://luukvdmeer.github.io/sfnetworks/articles/sfn02_preprocess_clean.html#smooth-pseudo-nodes
        convert(to_spatial_smooth, .clean = TRUE)
}
```

```{r}
cleaned <- clean(trimmed)
```

We calculate the weights (= distances) of the edges, will be used to determine the shortest paths between the two vertices

```{r}
calc_weights <- function(net){
    net |>
        activate("edges") |>
        mutate(weight = edge_length())
}
```

```{r}
network <- calc_weights(cleaned)
```

```{r}
# keep only the main connected component of the network
network <- network |>
    activate("nodes") |>
    filter(group_components() == 1)
```

Determine the corridor edge:

```{r}
get_target_points <- function(vertices, area, threshold = 0.0001){
    vertices |>
        st_as_sf() |>
        # keep threshold to check which points  intersect the polygons
        st_filter(area, .predicate = st_is_within_distance, dist = threshold) |>
        st_geometry()
}

get_corridor_edge <- function(network, area, vertices){
    # determine start and endpoint
    target_points <- get_target_points(vertices, area)


    # find shortest path
    paths <- st_network_paths(
        network,
        from = target_points[1],
        to = target_points[2],
        weights = "weight",
        type = "shortest"
    )

    edges <- network |> activate("edges") |> st_geometry()
    edge_path <- paths |> pull(edge_paths) |> unlist()
    edges[edge_path]
}
```

```{r}
corridor_edge_1 <- get_corridor_edge(network, area, vertices)
```

Visualize the computed corridor edge:

```{r}
leaflet() |>
    addTiles() |>
    addPolylines(data = corridor_edge_1 |> getGeomLatLon(), color = "blue")
```

We run the same steps on the "other side" of the river:

```{r}
area <- areas[2]

network <- trim(net, area, corridor_initial) |>
    clean() |>
    calc_weights() |>
    activate("nodes") |>
    filter(group_components() == 1)  # keep only the main connected component

corridor_edge_2 <- get_corridor_edge(network, area, vertices)
```

```{r}
leaflet() |>
    addTiles() |>
    addPolylines(data = corridor_edge_1 |> getGeomLatLon(), color = "blue") |>
    addPolylines(data = corridor_edge_2 |> getGeomLatLon(), color = "green") |>
    addPolylines(data = bbox |> getGeomLatLon(), color = "orange")
    # addPolylines(data = city_boundary |> getGeomLatLon(), color = "orange")
```

Finally, we combine the two corridor edges with the municipal boundary:

```{r}
corridor_edges <- st_union(corridor_edge_1, corridor_edge_2)

# corridor <- city_boundary |>
corridor <- bbox |>
    st_split(corridor_edges) |>
    st_collection_extract("POLYGON") |>
    st_as_sf() |>
    st_filter(waterway, .predicate = st_intersects)
```

```{r}
leaflet() |>
    addTiles() |>
    addPolylines(data = corridor |> getGeomLatLon(), color = "red")
```

Note that this is not ideal, as the municipal boundaries can be arbitrary and might exclude important end features of the corridors, so the user should have the option to input their own feature to cap the corridor ends. In the case of Bucharest, this can be the ring road.

```{r}
st_write(
    c(
        st_geometry(corridor),
        st_union(corridor_edge_1),
        st_union(corridor_edge_2)
    ),
    dsn = sprintf("data/generated/corridor_%s.gpkg", river_name),
    append=FALSE,
)
```
